Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2023-11-11 manuscript/manuscript.Rmd
Adding comments to manuscript text
Finished adding notes and easy edits to results section

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment.[@Karl1999; @Karl2017; @Karl2021] A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies. @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation, especially at the surface.

@@

# Results
## Metabolome changes significantly across the eddy dipole at DCM and 175m

We found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples while the surface was more homogenous (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the deep samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize this difference, we plotted an "average" metabolite per station at each depth by taking the median value of all the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was much more abundant than in any sample at this same depth in the rest of the dipole samples. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

@@

A large part of this metabolic response to SLA is due to the increased biomass in the cyclone DCM and the parallel increase in the anticyclone deep samples. We used a subset of the prior dataset, only those metabolites that we were able to quantify, and measured their in-situ concentrations in the particulate matter (Figure 2).

@@

*Figure 2: Concentrations of known compounds across the eddy dipole, broken down by depth. Stacked bars correspond to known compounds named in the legend underneath, with compounds not in the top 11 by absolute concentration grouped into the "Other" category. DMSP = dimethylsulfoniopropionate. Sea level anomaly (SLA) is shown by the red-blue color scale underneath the barplot with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone). The eddy centers (stations 6 and 12) have been labeled with arrows of the same color scheme in the final panel. Particulate carbon concentrations in µmol/L are shown as white dots and use the y-axis scale on the right side (200x the metabolite scale).*

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), representing an average concentration of 1.38 nM. At the surface, however, hydroxyisoleucine was the most concentrated metabolite at 1.53 nM due to its monotonic decrease with depth as opposed to a majority of the osmolytes (37/53) which reached maximum concentrations at the DCM instead (Figure 2).

Bulk particulate carbon (PC) measurements correlated well with overall metabolite concentrations ($r= .554$, p-value < 0.001) but this was largely driven by the lower biomass and metabolite abundance at depth. When analyzed at each depth separately, the DCM ($r_{DCM}= .790$, p-value = 0.003) and 175 meter samples ($r_{Deep}= .554$, p-value = 0.077) had weaker but still good correlations, but no such association was detected at the surface ($r_{Surface}= -.125$, p-value = 0.714).

@@

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.6 percentage points lower at 175 meters, p-value=0.002) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for SLA in the 175 meter samples if ~1.5x as many samples had been collected with mean and variances remaining similar.

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in both peak area and concentration. Heuristic exploration of nearby masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine (same mass as hydroxyisoleucine). This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining which compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time.

@@

@@

Finally, we performed a search of the untargeted data for homoserine betaine, a known *Trichodesmium* osmolyte [@Pade2016]. Although this compound was not in our standard library due to a lack of purchasable authentic standards, we identified three mass features in the 162.1125 *m/z* window with retention times of 7.1, 8.4, and 9.9 minutes. The feature at RT 9.9 matched our carnitine standard's retention time, leaving the two earlier peaks with putative identifications of threonine betaine at 7.1 and homoserine betaine at 8.4 given the elution order of threonine before homoserine in our authentic standards. This order is the reverse of that in @Pade2016, which showed homoserine betaine eluting prior to carnitine as expected given our use of a polar HILIC column and their use of a nonpolar Hypercarb column. They did not report an additional peak detected at this mass, leaving our putative identification of the threonine betaine as low-confidence. None of the three 162.1125 features showed the same degree of enrichment in MS11C215m_A as taurine or the other 13 compounds detected with that pattern. Although MS11C215m_A was the largest, second-largest, and third-largest peak for carnitine, threonine betaine, and homoserine betaine, respectively, it was never more than triple the average surface value or double the next-largest value.

@@

In addition to the absolute shifts in metabolite concentration due to biomass differences, we also explored the relative composition of the particulate matter.

@@

At the DCM, twenty-two compounds changed significantly in relative peak area across the eddy dipole. Seven of these were known metabolites, of which trigonelline and homarine were most enriched in the cyclone while arsenobetaine was the only metabolite with a larger peak area fraction in the anticyclone. Homarine showed a very large and highly significant shift, representing about 2.5% of the total peak area in the anticyclone but 7-8% in the cyclone center. Trigonelline, an isomer structurally very similar but biologically entirely distinct, had a surprisingly strong correlation with homarine (Pearson's $r=.855$) with peak areas consistently around one third of homarine's but large enough to have the second-largest shift in relative peak area of the significantly different compounds. Similarly, arsenobetaine represented about 0.1% of the peak area in the cyclone and 0.5-0.6% of the total peak area in the anticyclone. The most significantly different compounds at the DCM in each direction, however, were both unknowns. The lowest p-value ($6.4*10^{-6}$ after FDR correction) in the DCM data was a mass feature with *m/z* of 173.09211, a retention time also around 8.6 minutes, and enrichment in the anticyclone with a putative chemical formula of [M+H] = C$_7$H$_{13}$N$_2$O$_3$, possibly glycylproline or prolylglycine. The next-lowest (p = $2.5*10^{-4}$ after FDR correction) was enriched in the cyclone and had an *m/z* of 275.07119 and a retention time of 11.4 minutes with a putative formula of C$_{18}$H$_{11}$O$_3$.

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter difference in depth and the ~40 centimeter difference in SLA explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were well-correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3). A 2D NMDS plot of the samples showed good separation between the eddies, with cyclonic samples largely on the left side of Figure 3 while anticyclonic samples fell almost entirely to the right. Notably, the cyclonic samples show a much larger spread than the anticyclonic ones, indicating larger sample variability at the surface than the DCM. A depth gradient is visible along MDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom and left.

@@

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster 2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a neat DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

@@

Many of the known metabolites enriched in the cyclone are expected given the overall higher biomass, particularly as many have functions as osmolytes in the cell. Metabolites enriched in excess of biomass were taurine and isethionic acid, the reduced sulfur osmolytes dimethylsulfoniopropionate (DMSP) and dimethylsulfonioacetate (DMS-Ac), and the isomers homarine and trigonelline. A few metabolites more abundant in the anticyclone were also given putative identifications based on RT and *m/z* matching with internal standards later added. Of these, the putative arsenobetaine was the most significantly different among these with a peak area at the anticyclone DCM nearly quadruple that of the cyclone. Of note, the 173.09211 mass feature noted above was among the most significantly different between the two eddies, along with two isomers at an *m/z* of 170.11756 and retention times around 8-9 minutes that increased by a factor of 1.5 in the anticyclone (putative formula C$_7$H$_{14}$N$_4$O).

@@

As expected, TMAO was again one of the most abundant metabolites detected in the particulate matter with concentrations around 1.2 nM at the surface and 0.4 nM at the DCM, second only to the surprisingly high levels of the amino acid glycine (2 nM surface concentrations, 0.9 nM DCM concentrations), though the differences between the surface and DCM were not significant for either compound. We did detect a significant difference between depths for hydroxyisoleucine, once again expected, despite the aforementioned lower concentrations in this sample set (surface mean = 0.32 nM, DCM mean = 0.21 nM, t-test p-value=0.022 with n=12 samples at each depth).

>>>>

High frequency, long-term observations at Station ALOHA in the North Pacific Subtropical Gyre (NPSG) over the past 25 years have revealed temporal and spatial variability in what had prevously been considered a homogenous environment.[@Karl1999; @Karl2017; @Karl2021] A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the depth and intensity of the deep chlorophyll maximum (DCM) at the top of the nitricline differs dramatically between cyclonic and anticyclonic eddies. @Barone2022 reported a shift in the microbial community at the DCM with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies. @Hawco2021 and @Dugenne2023 showed that some of this difference was due to alterations in dissolved iron chemistry and nitrogen fixation, especially at the surface. [Anitra: so the surface is driving changes at the DCM?]

@@

# Results
## Metabolome variability across the eddy dipole

Using data for all molecular features, we found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the 175m samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

To characterize this difference, we plotted an "average" metabolite peak area per station at each depth by taking the median value of all the detected mass features after z-score normalization (Figure 1). As expected, the average metabolite collected at the exact center of the cyclonic eddy (Station 12) was much more abundant than in any sample at this same depth in the rest of the dipole samples. Similarly, the 175 meter samples showed the largest metabolite abundances in the anticyclone relative to other stations at this depth. This method also allowed us to see the remarkably sinusoidal response of the average metabolite across the stations. This response correlated negatively with SLA at the DCM (lower SLA = higher average values, $r_{DCM}= -.716$) but positively with SLA at the surface and at depth ($r_{surface} = .211$, $r_{deep} = .830$).

@@

We expected a large part of the difference in metabolomes with respect to SLA to be due to the increased biomass in the cyclone DCM and the parallel increase in the anticyclone deep samples. Within our molecular feature population, a subset are known metabolites that we quantified using standards (Figure 2).

@@

*Figure 2: Concentrations of known compounds across the eddy dipole, broken down by depth. Stacked bars correspond to known compounds, with the top 11 shown and all other metabolites summed in grey (n=[TBD]). DMSP = dimethylsulfoniopropionate. Sea level anomaly (SLA) is shown by the red-blue color scale underneath the barplot with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone). The eddy centers (stations 6 and 12) have been labeled with arrows of the same color scheme in the final panel. Particulate carbon concentrations in µmol/L are shown as white dots and use the y-axis scale on the right side (200x the metabolite scale) [Anitra: This implies that the metabolites are 1/200 of the POC. But you have plotted compound concentration not carbon for metabolites. Assuming each compound has an average of 4 carbon atoms, you would have to multiply by 4 so that is 1/50 (2%) or so of POC. I think you need to make this clear somehow. Or you could make an additional plot with C for each compound. I see supp fig 2 has the %POC so maybe that is enough but maybe call out that figure here.][Will: I agree, let's brainstorm].*

@@

We found that the overall concentration of known molecules was highest at the DCM of the cyclone, in line with our expectations and prior results. The most abundant quantified molecule overall was trimethylamine N-oxide (TMAO), with an average concentration of 1.38 nM. At the surface hydroxyisoleucine was the most abundant metabolite at 1.53 nM due to its monotonic decrease with depth as opposed to a majority of the osmolytes (37/53) which reached maximum concentrations at the DCM instead (Figure 2).

[Anitra: comments on the paragraph above: Maybe instead of this somewhat convoluted description you could just break it down by depth and say which compound was most abundant at each depth? Or which compound was most abundant in the cyclone and anticyclone. And then in a separate sentence just simply state that most compounds reached a ma at the DCM, but hydroxyisoleucine was an exception. Also, are you implying that hydroxyisoleucine is an osmolyte? Is that true? Also, what do you mean by "average"?]

Bulk particulate organic carbon (POC) measurements correlated well with overall [Anitra: what do you mean by overall here?] metabolite concentrations ($r= .554$, p-value < 0.001) but this was largely driven by both parameters having higher values in the surface than at depth. When analyzed at each depth individually, the DCM ($r_{DCM}= .790$, p-value = 0.003) and 175 meter samples ($r_{Deep}= .554$, p-value = 0.077) had weaker but still good correlations, but no such association was detected at the surface ($r_{Surface}= -.125$, p-value = 0.714).

@@

This correlation between PC [Anitra: Is there ATP data? Might be interesting to compare. Will: instinct says no, I think that ATP was really only collected during the diel phase and from a few spots during lagrangian. TODO: Double check the dashboard] and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. [Anitra: I do not see this at all. I see spikes in metabolite concentration that are not matched with spikes in PC. Mostly at the eddy centers. Do you not think this is suggestive of a change in the contribution of biomass to PC in the eddy centers? I would assume if you had biomass instead of PC you would be able to see biomass increase at the centers where metabolites increase. TODO: Check the same values using beam attenuation (interpolate if necessary)] Known metabolites contributed between approximately 1.5% and 5% of the carbon to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.6 percentage points lower at 175 meters, p-value=0.002) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). A higher proportion of PC in living biomass would translate to a higher proportion of metabolites in POC. This lack of difference may have been due to the small sample sizes (6 anticyclonic stations, 3 cyclonic, and 2 neither, each sampled in triplicate). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for SLA in the 175 meter samples if ~1.5x as many samples had been collected with mean and variances remaining similar.

[TODO: switch to POC instead of PC after confirming it's only the organics]

@@

One sample, taken from the surface at Station 11 (MS11C215m_A), was highly distinct from all others (including its paired triplicates) with several metabolites measuring two or more orders of magnitude larger in peak area and concentration both before and after normalization to internal standards. Heuristic exploration of potentially related masses identified a total of 14 compounds highly abundant in this one sample alone. The only identified compounds with this trend were taurine and acetyltaurine, though three others with *m/z* 146.1176, 148.0968, and 206.1388 were discovered to be likely isomers of known compounds (acetylcholine/(3-carboxypropyl)trimethylammonium, hydroxyisoleucine, dexpanthenol, respectively). One feature (*m/z* = 184.0637, RT = 7.0 minutes) was identified as choline sulfate using a second run of standards, while an additional mass feature at 168.0689 and RT = 8.5 was putatively flagged as taurine betaine (same mass as hydroxyisoleucine). This mass was also one of the most highly enriched in the MS11C215m_A sample after taurine itself, with a peak area nearly 30x that of the next largest sample. No clear explanation emerged explaining why some compounds were hyperabundant in this surface sample, though one possible explanation was the *Trichodesmium* bloom occurring at the same time.

@@

[TODO: Clean up this table. Make sure the RT values only 2-3 sigfigs, make column names legible, etc.]

@@

Finally, we performed a search of the untargeted data for homoserine betaine, a known *Trichodesmium* osmolyte [@Pade2016]. Although this compound was not in our standard library due to a lack of authentic standards, we identified three mass features in the 162.1125 *m/z* window with retention times of 7.1, 8.4, and 9.9 minutes. The feature at RT 9.9 matched our carnitine standard's retention time, leaving the two earlier peaks with putative identifications of threonine betaine at 7.1 and homoserine betaine at 8.4 given the elution order of threonine before homoserine in our authentic standards. This order is the reverse of that in @Pade2016, which showed homoserine betaine eluting prior to carnitine as expected given our use of a polar HILIC column and their use of a nonpolar Hypercarb column. They did not report an additional peak detected at this mass, leaving our putative identification of the threonine betaine as low-confidence. None of the three 162.1125 features showed the same degree of enrichment in MS11C215m_A as taurine or the other 13 compounds detected with that pattern. Although MS11C215m_A was the largest, second-largest, and third-largest peak for carnitine, threonine betaine, and homoserine betaine, respectively, it was never more than triple the average surface value or double the next-largest value.

@@

In addition to absolute shifts in metabolite concentration due to biomass differences, we also explored compositional shifts in the metabolomes.

@@

At the DCM, twenty-two compounds changed significantly in relative peak area across the eddy dipole. Seven of these were known metabolites, of which trigonelline and homarine were most enriched in the cyclone while arsenobetaine was the only metabolite with a larger fraction of the total peak area in the anticyclone. Homarine showed a very large and highly significant shift, representing about 2.5% of the total peak area in the anticyclone but 7-8% in the cyclone center. Trigonelline, an isomer structurally very similar but biologically entirely distinct, had a surprisingly strong correlation with homarine (Pearson's $r=.855$) with peak areas consistently around one third of homarine's but large enough to have the second-largest shift in relative peak area of the significantly different compounds. Similarly, arsenobetaine represented about 0.1% of the peak area in the cyclone and 0.5-0.6% of the total peak area in the anticyclone. The most significantly different compounds at the DCM in each direction, however, were both unknowns. The lowest p-value ($6.4*10^{-6}$ after FDR correction) in the DCM data was a mass feature with *m/z* of 173.09211, a retention time also around 8.6 minutes, and enrichment in the anticyclone with a putative chemical formula of [M+H] = C$_7$H$_{13}$N$_2$O$_3$, possibly glycylproline or prolylglycine. The next-lowest (p = $2.5*10^{-4}$ after FDR correction) was enriched in the cyclone and had an *m/z* of 275.07119 and a retention time of 11.4 minutes with a putative formula of C$_{18}$H$_{11}$O$_3$.

@@

To provide context for the sea level anomaly responses detected above, we additionally performed high-resolution depth sampling around the DCM at the center of both eddy poles. Here, we found that the ~40 meter difference in depth and the transition between the eddies explained similar amounts of variation in the data (R$^2_{depth}=0.253$, R$^2_{SLA}=0.238$) and were both highly significant factors (permutational p-values << 0.001). Additionally, both sample depth and SLA were well-correlated with the first principal component of the metabolite matrix (r$_{depth}=.766$, r$_{SLA}=.684$, fraction of variance explained by PC$_1$ = 30.8%) (Figure 3). A 2D NMDS plot of the samples showed good separation between the eddies, with cyclonic samples largely on the left side of Figure 3 while anticyclonic samples fell almost entirely to the right. Notably, the cyclonic samples show a much larger spread than the anticyclonic ones, indicating larger sample variability at the surface than the DCM. A depth gradient is visible along MDS 2, with the deepest samples generally at the top right of the plot and the shallowest ones closer to the bottom and left.

@@

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster 2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

@@

Many of the known metabolites enriched in the cyclone are expected given the overall higher biomass, particularly as many have functions as osmolytes in the cell. Metabolites enriched in excess of biomass were taurine and isethionic acid, the reduced sulfur osmolytes dimethylsulfoniopropionate (DMSP) and dimethylsulfonioacetate (DMS-Ac), and the isomers homarine and trigonelline. A few metabolites more abundant in the anticyclone were also given putative identifications based on RT and *m/z* matching with internal standards later added. Of these, the putative arsenobetaine was the most significantly different among these with a peak area at the anticyclone DCM nearly quadruple that of the cyclone. Of note, the 173.09211 *m/z* mass feature noted above was among the most significantly different between the two eddies, along with two isomers at an *m/z* of 170.11756 and retention times around 8-9 minutes that increased by a factor of 1.5 in the anticyclone (putative formula C$_7$H$_{14}$N$_4$O).

@@

As expected, TMAO was again one of the most abundant metabolites detected in the particulate matter with concentrations around 1.2 nM at the surface and 0.4 nM at the DCM, second only to the high levels of the amino acid glycine (2 nM surface, 0.9 nM DCM), though the differences between the surface and DCM were not significant for either compound. We did detect a significant difference between depths for hydroxyisoleucine, once again expected, despite the aforementioned lower concentrations in this sample set (surface mean = 0.32 nM, DCM mean = 0.21 nM, t-test p-value=0.022 with n=12 samples at each depth).
